Compiling Python Modules to Native Parallel Modules Using Pythran and OpenMP Annotations

### 2013-10-26 doc/papers/sc2013/paper.tex
SC2013: take into account reviewer3 comments

This paper presents the addition of OpenMP, a popular standard to turn
sequential C, C++ and FORTRAN application into parallel applications, to the
Pythran~\cite{pythran2013} compiler, a compiler that turn Python modules into
C++ meta-programs.  Section~\ref{sec:python-parallel} gives an overview of
existing approaches to compile Python functions or modules and shows a critical
lack of parallelism support. It also emphasises on the need for a
backward-compatible approach. Section~\ref{sec:python-static} presents the
Pythran compiler and the underlying runtime. Section~\ref{sec:python-openmp}
studies the validity of using OpenMP directives within the Python language in
% I would dispute the fact that OpenMP is "fine-grain", we're not talking about % vectorization here.
order to add fine-grain parallelism support to Python in the context of Pythran.
Finally, the Pythran compiler is benchmarked on several scientific kernels and
compared to Cython, Numba and Parakeet in Section~\ref{sec:validation}.

@@

There have also been several approaches to replace the GIL by Transactional
Memories~\cite{Riley2006,Tabba2010} but none of them made there way to the
mainstream interpreters. As a consequence, Python developers need to write
multi-threaded native modules in order to fully benefit from multiple cores.
This leads to a kind of computations referred as \emph{hybrid computations}.

@@

\subsection{Runtime Support}

% STL != standard library
Pythran runtime is based on the C++11 standard library which, unlike its C++03
version, is thread-safe. It implies that no data dependencies are added to the
original code.

A critical point of the design of the \texttt{pythonic++} runtime is memory
management, the very same aspect that led to the use of a GIL in CPython.
Memory management is implemented in Shed Skin through the general-purpose Boehm
garbage collector~\cite{boehm1991}, and Cython forbids usage of Python-managed
objects inside parallel regions, thus making memory management explicit for
these parts.

@@

Pythran's runtime is aware of these three kinds of iterators and supports
parallel iteration over random access iterators and iterators adapting random
ones.

@@

\subsubsection{Interaction With Pythran Runtime}

@@

We have used the Pythran tool described in the following % following??
 section to turn each
Python test function into a C++ function with the same directives and runtime
calls. Apart from the \texttt{threadprivate} directive and the
\texttt{collapse(\emph{n})} clause, all tests were successful.
\texttt{threadprivate} directives were held by global variables not supported in
Pythran yet~; and the C++ code generated by Pythran does not preserve the
perfect loop nesting required by the \texttt{collapse} clause.

@@

The target machine has  8 AMD Opteron 6176 SE cores and 8 GO of RAM. It is
running a Linux kernel 3.10-2-amd64, x86\_64 GNU/Linux. The C++ compiler used is
\texttt{g++} version 4.8.1 and all applications are compiled using the
\texttt{-Ofast} flag. Pythran generated programs are linked with Boost.Python
version 1.54. The Pythran version is extracted from the \texttt{sc2013} branch
of the git repository \url{https://github.com/serge-sans-paille/pythran}. Python
programs are run using the CPython implementation, version 2.7.5. Cython
programs are generated using version 0.19.1. Numba and Parakeet are installed
from the \texttt{master} branch of their respective repositories.
% dommage pour numba et parakeet qui sont les seuls "mal spécifiés"...

@@

    \centering
    \begin{tabular}{|l|r|}
        \hline
        Language & SLOC     \\
        \hline
        C       & 102       \\
        Python  & 30        \\
        Pythran & 30        \\
        Pythran+OMP    & 31 \\
        \hline
    \end{tabular}

@@

Figure~\ref{fig:pb-rosen} shows the execution time of the computation of one
step of the GrowCut algorithm used in image processing. On this example,
Parakeet performs better than Cython and Pytrhan and way better than Numba, but
paralleizing the outermost loop with OpenMP gives a $\times 2.65$ speedup to
Pythran that helps it outperform Parakeet.

@@

To achieve this goal, we designed Pythran, a translator from a subset of the
Python language to C++. Pythran turns regular Python modules annotated with
OpenMP directives and a few type annotations into native parallel module. The
input module remains compatible with the standard interpreter and the underlying
runtime is compatible with parallel constructs. While Pythran accepts a limited
input, we extend it progressively and with a great care on its stability. As
such, even if its development is community-based, it is more than a research
prototype and should be usable on real world code.

@@

This work has received fundings from \textsc{silkan} and the French ANR through
the CARP project. The authors thank Adrien \textsc{Merlini} and Alan
\textsc{Raynaud} for their work on the runtime, Albert \textsc{Cohen},
B{\'e}atrice \textsc{Creusillet}, Fabien \textsc{Dagnat}, Fran{\,c}ois
\textsc{Irigoin} and Ronan \textsc{Keryell} for their valuable reviews.

>>>>

This paper presents the addition of OpenMP, a popular standard to turn
sequential C, C++ and FORTRAN application into parallel applications, to the
Pythran~\cite{pythran2013} compiler, a compiler that turns Python modules into
C++ meta-programs.  Section~\ref{sec:python-parallel} gives an overview of
existing approaches to compile Python functions or modules and shows a critical
lack of parallelism support. It also emphasises on the need for a
backward-compatible approach. Section~\ref{sec:python-static} presents the
Pythran compiler and the underlying runtime library.
Section~\ref{sec:python-openmp} studies the validity of using OpenMP directives
within the Python language in
% I would dispute the fact that OpenMP is "fine-grain", we're not talking about % vectorization here.
order to add fine-grain parallelism support to Python in the context of Pythran.
Finally, the Pythran compiler is benchmarked on several scientific kernels and
compared to Cython, Numba and Parakeet in Section~\ref{sec:validation}.

@@

There have also been several approaches to replace the GIL by Transactional
Memories~\cite{Riley2006,Tabba2010} but none of them made their way to the
mainstream interpreters. As a consequence, Python developers need to write
multi-threaded native modules in order to fully benefit from multiple cores.
This leads to a kind of computations referred as \emph{hybrid computations}.

@@

\subsection{Runtime Library Support}

% STL != standard library
Pythran runtime library is based on the C++11 standard library which, unlike its
C++03 counterpart, is thread-safe. It implies that no data dependencies are
added to the original code.

A critical point of the design of the \texttt{pythonic++} runtime library is
memory management, the very same aspect that led to the use of a GIL in CPython.
Memory management is implemented in Shed Skin through the general-purpose Boehm
garbage collector~\cite{boehm1991}, and Cython forbids usage of Python-managed
objects inside parallel regions, thus making memory management explicit for
these parts.

@@

Pythran's runtime library is aware of these three kinds of iterators and
supports parallel iteration over random access iterators and iterators adapting
random ones.

@@

\subsubsection{Interaction With Pythran Runtime Library}

@@

We have used the Pythran tool described in the following % following??
 section to turn each
Python test function into a C++ function with the same directives and runtime
library calls. Apart from the \texttt{threadprivate} directive and the
\texttt{collapse(\emph{n})} clause, all tests were successful.
\texttt{threadprivate} directives were held by global variables not supported in
Pythran yet~; and the C++ code generated by Pythran does not preserve the
perfect loop nesting required by the \texttt{collapse} clause.

@@

The target machine has  8 AMD Opteron 6176 SE cores and 8 GB of RAM. It is
running a Linux kernel 3.10-2-amd64, x86\_64 GNU/Linux. The C++ compiler used is
\texttt{g++} version 4.8.1 and all applications are compiled using the
\texttt{-Ofast} flag. Pythran generated programs are linked with Boost.Python
version 1.54. The Pythran version is extracted from the \texttt{sc2013} branch
of the git repository \url{https://github.com/serge-sans-paille/pythran}. Python
programs are run using the CPython implementation, version 2.7.5. Cython
programs are generated using version 0.19.1. Numba and Parakeet are installed
from the \texttt{master} branch of their respective repositories.
% dommage pour numba et parakeet qui sont les seuls "mal spécifiés"...

@@

    \centering
    \begin{tabular}{|l|c|}
        \hline
        Language & Source Lines Of Code (SLOC) \\
        \hline
        C       & 102       \\
        Python  & 30        \\
        Pythran & 30        \\
        Pythran+OMP    & 31 \\
        \hline
    \end{tabular}

@@

Figure~\ref{fig:pb-growcut} shows the execution time of the computation of one
step of the GrowCut algorithm used in image processing. On this example,
Parakeet performs better than Cython and Pytrhan and way better than Numba, but
paralleizing the outermost loop with OpenMP gives a $\times 2.65$ speedup to
Pythran that helps it outperform Parakeet.

@@

To achieve this goal, we designed Pythran, a translator from a subset of the
Python language to C++. Pythran turns regular Python modules annotated with
OpenMP directives and a few type annotations into native parallel module. The
input module remains compatible with the standard interpreter and the underlying
runtime library is compatible with parallel constructs. While Pythran accepts a
limited input, we extend it progressively and with a great care on its
stability. As such, even if its development is community-based, it is more than
a research prototype and should be usable on real world code.

@@

This work has received fundings from \textsc{silkan} and the French ANR through
the CARP project. The authors thank Adrien \textsc{Merlini} and Alan
\textsc{Raynaud} for their work on the runtime library, Albert \textsc{Cohen},
B{\'e}atrice \textsc{Creusillet}, Fabien \textsc{Dagnat}, Fran{\,c}ois
\textsc{Irigoin} and Ronan \textsc{Keryell} for their valuable reviews.
